---
title: "QSAR with ChEMBL data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## QSAR

Here, I will build a quantitative structure activity relationship model to predict chemical bioactivity for a protein of interest. The data will come from the ChEMBL database, an SQL database that contains bioactivities for many different proteins. The ChEMBL database must be downloaded from: https://www.ebi.ac.uk/chembl/  I will make a QSAR model for the BRAF cancer target. 

```{r cars}
library(dplyr)
library(caret)
library(xgboost)
require(rcdk)
require(Rcpi)

path_to_chembl <- "/home/kb/Desktop/chembl/chembl_22.db"
target <- "P15056"

chembl_db <- src_sqlite(path_to_chembl, create = TRUE)

```


Now I will extract the relevant data from the SQL database.

```{r}
#establish chembl data frames
activities <- tbl(chembl_db, "activities")
molecule_dictionary <- tbl(chembl_db, "molecule_dictionary")
assays <- tbl(chembl_db, "assays")
component_sequences <- tbl(chembl_db, "component_sequences")
target_components <- tbl(chembl_db, "target_components")
compound_structures <- tbl(chembl_db, "compound_structures")

#get bioactivities for target
component <- component_sequences %>% filter(accession == target)
component <- collect(component, n = Inf)
tid <- target_components %>% filter(component_id == component$component_id)
tid <- collect(tid, n = Inf)
assay_ids <- assays %>% filter(tid %in% tid$tid & confidence_score > 7)
assay_ids <- collect(assay_ids , n = Inf)
bioactivities <- activities %>% filter(assay_id %in% assay_ids$assay_id & standard_units == "nM" & standard_value > 0) %>%
  select(molregno, standard_value)
bioactivities <- collect(bioactivities, n = Inf)
bioactivities <- bioactivities %>% filter(standard_value > 0) %>% mutate(pIC50 = -log10(standard_value * 10^(-9)))
molecules <- compound_structures %>% select(molregno, canonical_smiles) %>% filter(molregno %in% bioactivities$molregno)
molecules <- collect(molecules, n = Inf)
bioactivities <- left_join(bioactivities, molecules, by = "molregno")
```

Next, I will calculate the mean bioactivity for each compound, and convert the compounds to circular fingerprints with the circularFP function. Then I split the data into training and test sets
```{r}
#find mean bioactivity for each molecule, then remove those containers that are NA
bioactivities <- bioactivities %>% group_by(molregno, canonical_smiles) %>% 
  summarize(averagepIC50 = mean(pIC50)) %>% ungroup()
containersTrain <- parse.smiles(bioactivities$canonical_smiles)
bioactivities <- bioactivities[!is.na(containersTrain),]
containersTrain <- containersTrain[!is.na(containersTrain)]


circularFP <- function(cmpdlist){
  fptest <- get.fingerprint(cmpdlist[[1]], type = "circular", fp.mode = 'bit')
  fpmatrix <- matrix(data = 0, nrow = length(cmpdlist), ncol = attr(fptest, "nbit"))
  for(i in 1:length(cmpdlist)){
    fp <- get.fingerprint(cmpdlist[[i]], type = "circular", fp.mode = 'bit')
    fpmatrix[i, attr(fp, "bits")] <- 1
  }
  lostColumns <- nearZeroVar(fpmatrix)
  fpmatrix <- fpmatrix[, -nearZeroVar(fpmatrix)]
  attr(fpmatrix, "LostColumns") <- lostColumns
  
  return(fpmatrix)
}

fingerprints <- circularFP(containersTrain)
nrow(fingerprints) == length(containersTrain)

smp_size <- floor(0.75 * nrow(bioactivities))

## set the seed to make your partition reproductible
set.seed(123)
trainIndex <- sample(seq_len(nrow(bioactivities)), size = smp_size)
x.tr <- fingerprints[trainIndex, ]
x.te <- fingerprints[-trainIndex, ]
y.tr <- bioactivities$averagepIC50[trainIndex]
y.te <- bioactivities$averagepIC50[-trainIndex]
```

Now the model is trained with caret, using "xgbTree". And then we look at the diagnostic plot....
```{r}
ctrl <- trainControl(method = 'repeatedcv', number = 5, repeats = 2, summaryFunction = defaultSummary)
caret_model <- train(x.tr, y.tr, method = "xgbTree", trControl = ctrl, tuneLength =   5, metric = 'RMSE')
plot(caret_model)
plot(y.te, predict(caret_model, x.te), xlim = range(y.te), 
     ylim = range(y.te), 
     main = paste("Diagnostic plot for: ", target), xlab = "Data", ylab = "Predicted", cex = 0.2)
abline(a = 0, b = 1)
```


It looks pretty good. RMSE is 0.7. These models are very useful when trying to build polypharmacology in to a drug.